---
title: "hw1"
author: "yuner rao"
format:
  html:
    self-contained: true
    code-fold: true
editor: visual
---

```{r}
#1.1 importing dataset
soybean <- read.csv("data.csv")

names(soybean) <- c(
  "Season",
  "Cultivar",
  "Repetition",
  "PH",
  "IFP",
  "NLP",
  "NGP",
  "NGL",
  "NS",
  "MHG",
  "GY"
)

head(soybean)
str(soybean)
dim(soybean)
```

### Interpretation1.1

This soybean dataset includes 320 records and 11 columns. `Season`, `Cultivar`, and `Repetition` are grouping factors for the field experiment. The other variables are numeric agronomic traits. The `str()` function confirms variable types, and `head()` shows the renamed columns are loaded correctly.

```{r}
#problem 1.2 verify dataset
nrow(soybean)
length(unique(soybean$Cultivar))
unique(soybean$Season)
length(unique(soybean$Season))
season_counts <- table(soybean$Season)

season_counts
```

### Interpretation1.2

The code verifies there are 320 observations and 40 unique cultivars in the soybean dataset. Data was collected from two seasons. The table of counts shows an equal number of 160 observations for each season, forming a balanced experimental design.

```{r}
#problrm 1.3.1 calculate Pearson correlation coefficient
cor(soybean$PH, soybean$GY)
```

### Interpretation1.3.1

the Pearson correlation between plant height (PH) and (GY) is **0.1232807**. This indicates a positive liner association between plant height and grain yield.

```{r}
#problrm 1.3.2 calculate Pearson correlation for each season and compare the values
cor_season_1 <- with(
  soybean[soybean$Season == 1, ],
  cor(PH, GY)
)

cor_season_2 <- with(
  soybean[soybean$Season == 2, ],
  cor(PH, GY)
)

cor_season_1
cor_season_2
if (cor_season_1 > cor_season_2) {
  "Season 1 has the higher correlation."
} else {
  "Season 2 has the higher correlation."
}
```

### Interpretation1.3.2

Compared with Season2, **Season 1 a has the higher correlation**.

```{r}
#problem 1.3.3 calculate the GY of the soybean observation with the greatest height
soybean[which.max(soybean$PH), ]
```

### Interpretation1.3.3

The greatest plant height is 94.8 cm. This observation corresponds to cultivar FTR 4288 IPRO in Season 2 (Repetition 3), with a grain yield of approximately **3304.68**.

```{r}
#the percentage of observations
mean_MHG <- mean(soybean$MHG)
mean_MHG
mean_MHG <- mean(soybean$MHG)

percentage <- mean(soybean$MHG > mean_MHG) * 100

percentage
```

### Interpretation1.3.4

**47.1875%** Observations have a thousand seed weight greater than the overall average thousand seed weight.

```{r}
# problem 1.3.5 who has highest GY
cultivar_GY <- aggregate(
  GY ~ Cultivar,
  data = soybean,
  FUN = mean
)

cultivar_GY[
  which.max(cultivar_GY$GY),
]
```

### Interpretation1.3.5

**98R30CE** has the highest average GY

```{r}
#problem 1.4 establish a table showing the two average values
quant_vars <- c(
  "PH", "IFP", "NLP", "NGP",
  "NGL", "NS", "MHG", "GY"
)

overall_means <- colMeans(
  soybean[, quant_vars]
)

overall_means
season_means <- aggregate(
  soybean[, quant_vars],
  by = list(Season = soybean$Season),
  FUN = mean
)

season_means
overall <- data.frame(
  Season = "Overall",
  t(overall_means),
  check.names = FALSE
)
```

```{r}
#problem 1.5 carry out a series of t-tests 
```

### Interpretation1.5

Null and alternative hypotheses.

```         
$H_0: \mu_1 = \mu_2$
$H_A: \mu_1 \neq \mu_2$
```

```{r}
#t-test(Welch)
t_test_result <- t.test(
  GY ~ Season,
  data = soybean
)

t_test_result
```

```{r}
#t-test(equal-variance)
t.test(
  GY ~ Season,
  data = soybean,
  var.equal = TRUE
)
```

I used Welch's two-sample t-test as the primary analysis because it does not require equal population variances. I also checked the equal-variance version as a sensitivity analysis.Both t-tests lead to the same conclusion. Although the observed mean grain yield was slightly higher in Season 1 (3428.452) than in Season 2 (3408.655), **the difference was not statistically significant at the 0.05 significance level**. Therefore, based on these t-tests, we do not have sufficient evidence that mean grain yield differs between the two seasons.

```{r}
#problem 2.1 import dataset
adult <- read.csv(
  "adult.data",
  header = FALSE,
  na.strings = "?",
  strip.white = TRUE
)

names(adult) <- c(
  "age",
  "workclass",
  "fnlwgt",
  "education",
  "education_num",
  "marital_status",
  "occupation",
  "relationship",
  "race",
  "sex",
  "capital_gain",
  "capital_loss",
  "hours_per_week",
  "native_country",
  "income"
)
dim(adult)
```

### Interpretation2.1

By importing dataset, we know that there are **32561 observations** and **15 variables.**

```{r}
#problem 2.2 clean and simplify the variable names
names(adult)
```

### Interpretation2.2

The variable names were cleaned by replacing hyphens with underscores and using concise, descriptive names that are easier to work with in R.

```{r}
#problem 2.3 examine missing values
colSums(is.na(adult))
```

### Interpretation2.3

Missing values are present in three variables: workclass (1,836), occupation (1,843), and native country (583). All other variables have no missing observations.

```{r}
#problem 2.4 choose restricted samples
nrow(adult)#before filtering
adult_hours <- subset(adult, hours_per_week >= 10)
nrow(adult_hours)#after filtering
```

### Interpretation2.4

Before filtering:32561

After filtering:32103

```{r}
#problem 2.5 remove observations
adult_clean <- subset(
  adult_hours,
  !is.na(occupation) &
    !is.na(workclass) &
    !is.na(native_country)
)
nrow(adult_clean)
colSums(is.na(adult_clean))
```

### Interpretation2.5

These observations were removed because occupation, workclass, and native country are required variables, and missing values would prevent us from using these observations reliably in the analysis.

```{r}
#problem 2.6 remove selected values
adult_clean <- subset(
  adult_clean,
  age >= 18 &
    age <= 90 &
    education != "Preschool"
)
nrow(adult_clean)
```

### Interpretation2.6

Observations with age below 18, age above 90, or education level equal to Preschool were removed. The resulting final sample contains **29523** observations.

```{r}
#problem 2.7 final sample size
cat("Final sample size:", nrow(adult_clean), "\n")
```

### Interpretation2.7

Final sample size is **29523**

```{r}
#problem 3
nextHappy <- function(n) {
  
  # Check that n is a single positive integer
  if (length(n) != 1 || !is.numeric(n) ||
      !is.finite(n) || n <= 0 || n != floor(n)) {
    stop("n must be a single positive integer.")
  }
  
  # Convert the number to individual digits
  digits <- as.integer(strsplit(as.character(n), "")[[1]])
  
  # Return the sum of squared digits
  return(sum(digits^2))
}
nextHappy(19)
nextHappy(82)


#sequence

happySequence <- function(n) {
  
  # Check that n is a single positive integer
  if (length(n) != 1 || !is.numeric(n) ||
      !is.finite(n) || n <= 0 || n != floor(n)) {
    stop("n must be a single positive integer.")
  }
  
  # Store the sequence
  sequence <- n
  
  # Keep track of values that have already appeared
  seen <- n
  
  while (n != 1) {
    
    # Compute the next value
    n <- nextHappy(n)
    
    # Check whether a cycle has occurred
    if (n %in% seen) {
      return(list(
        sequence = c(sequence, n),
        isHappy = FALSE
      ))
    }
    
    # Add the new value to the sequence
    sequence <- c(sequence, n)
    seen <- c(seen, n)
  }
  
  # The sequence reached 1
  return(list(
    sequence = sequence,
    isHappy = TRUE
  ))
}
happySequence(19)
happySequence(82)
```

```{r}
results <- lapply(1:1000, happySequence)

is_happy <- sapply(
  results,
  function(x) x$isHappy
)

# 1. Number of Happy Numbers
num_happy <- sum(is_happy)
num_happy

# 2. Percentage of Happy Numbers
percent_happy <- mean(is_happy) * 100
percent_happy

# 3. Longest Happy sequence
sequence_lengths <- sapply(
  results,
  function(x) {
    if (x$isHappy) length(x$sequence) else NA
  }
)

max_length <- max(sequence_lengths, na.rm = TRUE)

longest_values <- which(
  sequence_lengths == max_length
)

max_length
longest_values
```
